---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## 1. Loading and preprocessing the data

- Activity data is downloaded as a zip file from the following source:

"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

``` {r echo = TRUE}
setwd("/Users/kumarnv/Rprojects/reproducibledata/project1")
activityUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

setwd("/Users/kumarnv/Rprojects/reproducibledata/project1")
zipFile <- "activity.zip"
activityFile <- "./activity.csv"

if (!file.exists("./data")) {
  dir.create("./data")
}

if (!file.exists(zipFile)) {
  download.file(activityUrl, destfile = zipFile, method = "curl")
}
activityData <- read.csv(activityFile)
```

## 2. Histogram of  total number of steps taken each day
```{r echo = TRUE}

stepsDaily <- aggregate(. ~ date, data = activityData, sum)$steps
hist(stepsDaily, breaks = 10)

```


## 3. What is mean total number of steps taken per day?
```{r echo = TRUE}
print(paste("daily steps mean", round(mean(stepsDaily))))
print(paste("daily steps median", round(median(stepsDaily))))
```


## 4. What is the average daily activity pattern?

```{r echo= TRUE}

plot(stepsDaily, type ="l", xlab = "Day#", main = "Daily activity pattern")

```
```{r}
stepsByInterval <- aggregate(steps ~ interval, activityData, mean)
plot(stepsByInterval, type="l", main= "Daily steps pattern by interval", ylab = "steps  mean")
```
## 5-miniute interval with max number of steps- without imputation
```{r}

activityMax <- max(stepsByInterval$steps)
# interval corresponding to the max number of steps
intervalMax <- stepsByInterval[stepsByInterval$steps == activityMax, 1]
print(paste("Interval corresponding to maximum activity", intervalMax))
```

## 6. Imputing missing values

- Used Misc package to impute missing values with mean number of steps

```{r, echo=TRUE}

#install.packages("Hmisc")
library(Hmisc)


totalNa <- sum(!complete.cases(activityData))
totalNa
activityDataImp <- activityData
activityDataImp$steps <- with(activityData, impute(steps, mean))
activityDataImp$steps <- round(activityDataImp$steps)
stepsDailyImp <- aggregate(. ~ date, data = activityDataImp, sum)$steps
print(paste("daily mean", round(mean(stepsDailyImp))))
#mean(stepsDailyImp)
#median((stepsDailyImp))
#plot(stepsDailyImp, type = "l", xlab = "day#", main = "Daily activity after data imputation")
```
## 7. Histogram of the total number of steps taken daily after data imputation. 

```{r echo=TRUE}
hist(stepsDailyImp, breaks = 10, main = "daily steps after data imputation", xlab = "Number of steps taken daily")
```

## 8. Are there differences in activity patterns between weekdays and weekends?

```{r echo = TRUE, fig.height= 6}
par(mfrow = c(2,1))
hist(stepsDaily, breaks = 10, main = "daily steps before data imputation", xlab = "Number of steps taken daily")
hist(stepsDailyImp, breaks = 10, main = "daily steps after data imputation", xlab = "Number of steps taken daily")
```
- Differences between weekday and weekends aren't obvious from histograms

## 8. Panel plot comparing average number of steps taken per 5-minute interval.

```{r fig.height=4}

par(mfrow = c(2,1))
library(dplyr)
library(lattice)
# Impute steps in activitydata
activityDataImp <- activityData
activityDataImp$steps <- with(activityData, impute(steps, mean))
activityDataImp$steps <- round(activityDataImp$steps)
imputedData <- mutate(activityDataImp, weekdays(as.Date(activityDataImp$date)))
colnames(imputedData)[4] <- "day"

imputedData$Daytype <- factor(imputedData$day, levels = c("Monday", "Tuesday","Wednesday", "Thursday", "Friday","Saturday","Sunday"), labels = c("weekday", "weekday", "weekday", "weekday", "weekday", "weekend","weekend"))
aggregateImputedData <- aggregate(steps~interval+Daytype, imputedData, mean)
xyplot(steps~interval|Daytype,
       data=aggregateImputedData,
       main="Interval Averages, Weekday v. Weekend",
       xlab="Interval",
       layout=c(1,2),
       type="l")


```
- daily activity pattern on weekdays appears to be different  from that on weekends.
